

Software engineering is a discipline that integrates process, methods, and tools for the development of computer software. To support it, Application lifecycle management systems are software tools that assist in the organization and moderation of a project throughout its life cycle. There is an infinite number of possibilities to define a project life cycle, and the needed tools to help it completion. Consequently, a huge variety of management systems are available on today's market, specifically designed to conform to a diverse number of management methodologies. Nevertheless, most are proprietary, very specialized, with little to no customization, making very hard to find one that perfectly fit one's needs.

A different type of software systems development is Software Product Line Engineering -- SPLE. SPL is a methodology for developing a diversity of related software products and software-intensive systems. During the development of a SPL, a wide range of artifacts needs to be created and maintained to preserve the consistency of the family model during development, and it is important to manage the SPL variability and the traceability among those artifacts. However, this is a hard task, due to the heterogeneity of assets developed during product line engineering. Maintaining the traceability and artifacts updated manually is error-prone, time consuming and complex. Utilizing a tool for supporting those activities is essential. 

In this work, we propose the Software Product Line Integrated Construction Environment (SPLICE). That is a web-based life cycle management tool for managing, in an automated way, the software product line activities. This initiative intends to support most of the SPL process activities such as scoping, requirements, architecture, testing, version control, evolution, management and agile practices. This was archived with the integration of a framework around an established tool, providing an easy way for handling the usage of different metamodels.

We present a lightweight metamodel which integrates the processes of the SPL lifecycle agile practices, the implementation of a tool that uses the proposed metamodel, and a case study that reflect the feasibility and flexibility of this solution especially for different scenarios and processes



\begin{keywords}
software product line, agile, SPL, Application lifecycle management , tool, metamodel
\end{keywords}